---
title: Simple single-cell analyses
author:
- name: Aaron Lun
  email: infinite.monkeys.with.keyboards@gmail.com
date: "Revised: March 11, 2026"  
output: BiocStyle::html_document
package: augere.solo
vignette: >
  %\VignetteIndexEntry{Simple single-cell analyses}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo=FALSE, results="hide"}
knitr::opts_chunk$set(error=FALSE, message=FALSE, warning=FALSE)
self <- BiocStyle::Biocpkg("augere.solo")
```

# Overview

`r self` implements [**augere** pipelines](https://github.com/augere-bioinfo) for a variety of simple single-cell analyses.
This use `r BiocStyle::Biocpkg("scrapper")` for routine steps such as quality control, clustering and marker detection.
We also use `r BiocStyle::Biocpkg("SingleR")` for automatic cell type annotation based on pre-labelled references.
Each pipeline creates a self-contained Rmarkdown report that documents each step in the analysis.

# Quick start

To install this package, follow the usual [instructions for Bioconductor](https://bioconductor.org/install):

```r
install.packages("BiocManager") # if not already available
BiocManager::install("augere.solo")
```

Let's load up our favorite example dataset from the `r BiocStyle::Biocpkg("scRNAseq")` package:

```{r}
library(scRNAseq)
sce.z <- ZeiselBrainData()
sce.z
```

Now we call `runSolo()` to perform a routine analysis of this dataset:

```{r, fig.keep="none"}
library(augere.solo)
outdir.z <- tempfile()
res.z <- runSolo(sce.z, qc.mito.regex = "^mt-", output.dir = outdir.z)
```

This returns a list of various results, including QC metrics, marker rankings and a `SingleCellExperiment` decorated with clusterings and reduced dimensions: 

```{r}
# Checking out the QC statistics.
res.z$qc.rna

# Checking out the reduced dimensions.
library(scater)
plotReducedDim(res.z$sce, "TSNE", colour_by = "graph.cluster")

# Checking out the marker rankings for each cluster.
# We use previewMarkers() just for a prettier print.
scrapper::previewMarkers(res.z$markers.rna[["1"]])
```

The function generates an Rmarkdown report containing (almost) all of the steps required to reproduce the analysis.
Let's have a look at it:

```{r}
fname <- file.path(outdir.z, "report.Rmd")
cat(head(readLines(fname), 50), sep="\n")
```

The function will also save the results to file inside a `results` subdirectory, which can be loaded back into the session using the `readResult()` function:

```{r}
reloaded <- readResult(file.path(outdir.z, "results", "markers-rna", "2"))
scrapper::previewMarkers(reloaded$x) # preview of the marker rankings
str(reloaded$metadata) # along with some metadata
```

# Multi-batch analyses

Larger datasets often have uninteresting variation associated with a blocking factor, e.g., batches.
To illustrate, let's combine the Zeisel dataset with another brain dataset:

```{r}
sce.t <- TasicBrainData()
common.brain <- intersect(rownames(sce.z), rownames(sce.t))
combined.brain <- combineCols(
    sce.z[common.brain,],
    sce.t[common.brain,]
)
combined.brain$study <- rep(c("zeisel", "tasic"), c(ncol(sce.z), ncol(sce.t)))
combined.brain
```

We now instruct `runSolo()` to block on the study of origin for each cell.
In many steps (e.g., variance calculations, differential expression), this will force all calculations to be performed within each level of the blocking factor.
This avoids interference from uninteresting differences between studies.

```{r, fig.keep="none"}
outdir.com <- tempfile()
res.com <- runSolo(
    combined.brain,
    block = "study",
    qc.mito.regex = "^mt-",
    output.dir = outdir.com,
    save.results = FALSE # skipping to save time for this vignette build.
)
```

We will also apply mutual nearest neighbors (MNN) correction to remove differences between studies in the low-dimensional space;
the MNN-corrected coordinates are then used for all downstream analyses.

```{r}
# Now we have an extra set of MNN-corrected coordinates. 
reducedDimNames(res.com$sce)

# Examining the distribution of clusters between studies.
table(res.com$sce$graph.cluster, res.com$sce$study)

# Visualizing the distribution in the UMAP space.
plotReducedDim(res.com$sce, "UMAP", colour_by = "study")
```

# With proteomics

`runSolo()` can also handle proteomics data from CITE-seq experiments.
We'll demonstrate on yet another example dataset, where the ADT counts are held in an alternative experiment:

```{r}
sce.k <- KotliarovPBMCData()
sce.k <- sce.k[,1:5000] # to cut build time for this vignette! 
sce.k
```

We specify the experiment containing the ADT counts in `runSolo()`.

```{r, fig.keep="none"}
outdir.k <- tempfile()
res.k <- runSolo(
    sce.k,
    qc.mito.regex = "^MT-",
    adt.experiment = "ADT",
    output.dir = outdir.k,
    save.results = FALSE # skipping to save time for this vignette build.
)
```

This applies quality control, normalization and dimensionality reduction on the ADT data,
and then combines the principal components with those from RNA into a single embedding for downstream analyses like clustering and visualization.
The goal is to incorporate information from both modalities in the distance calculations,
such that significant differences in either modality is capable of creating a new cluster.

```{r}
# Now we have an extra set of combined coordinates. 
reducedDimNames(res.k$sce)

# Checking out the ADT-specific QC metrics.
res.k$qc.adt

# Checking out the reduced dimensions.
plotReducedDim(res.k$sce, "TSNE", colour_by = "graph.cluster")

# We can also look at the ADT-specific markers for a cluster.
scrapper::previewMarkers(res.k$markers.adt[[1]])
```

We could also analyze the ADT modality by itself by setting `rna.experiment=NULL`.

# Cell type annotation

`r self` implements a wrapper around `r BiocStyle::Biocpkg("SingleR")` for cell type annotation.
We can either use one of the pre-defined references from the `r BiocStyle::Biocpkg("celldex")` package:

```{r, fig.keep="none"}
outdir.celldex <- tempfile()
pred.celldex <- runAnnotate(
    sce.z,
    configureReferenceAnnotation("MouseRNAseqData", "label.main"),
    output.dir = outdir.celldex
)
table(pred.celldex$predictions[[1]]$labels)
```

Or we can use another single-cell dataset.
Here, we set `ref.aggregate=TRUE` to speed up the classification by aggregating single-cells in the reference into a smaller number of representative clusters.

```{r, fig.keep="none"}
outdir.tasic <- tempfile()
pred.tasic <- runAnnotate(
    sce.z,
    configureReferenceAnnotation(sce.t, "primary_type", ref.assay="counts", ref.aggregate=TRUE),
    output.dir = outdir.tasic
)
table(pred.tasic$predictions[[1]]$labels)
```

We can even pass multiple references, in which case `runAnnotate()` will annotate against each reference separately and then choose between the best labels over all references.

```{r, fig.keep="none"}
outdir.multi <- tempfile()
pred.multi <- runAnnotate(
    sce.z,
    list(
        rnaseq=configureReferenceAnnotation("MouseRNAseqData", "label.main"),
        immgen=configureReferenceAnnotation("ImmGenData", "label.main")
    ),
    output.dir = outdir.multi
)

# Results for annotation against the individual references:
names(pred.multi$predictions)

# As well as the combined results:
table(pred.multi$combined$labels, pred.multi$combined$reference)
```

# Session information {-}

```{r}
sessionInfo()
```
