---
title: "Automatic annotation of scRNA-seq data"
author: <%= AUTHOR %>
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc_float: yes
    titlecaps: false 
  md_document:
    preserve_yaml: yes
---

```{r preamble, echo=FALSE}
knitr::opts_chunk$set(error=FALSE, warning=FALSE, message=FALSE)
```

# Test data loading

We load the `SingleCellExperiment` object corresponding to our test dataset.
Each row corresponds to a gene while each column corresponds to a cell.

```{r}
:BEGIN test-setup 
:END
test
```

:BEGIN replace-test-rownames
We replace the row names to obtain gene identifiers that are consistent with those used in the reference dataset.

```{r}
new.test.rownames <- SummarizedExperiment::rowData(test)[,<%= ALTERNATIVE_NAMES %>]
new.test.rownames[is.na(new.test.rownames)] <- ""
old.test.rownames <- rownames(test)
rownames(test) <- new.test.rownames
head(new.test.rownames)
```
:END

:BEGIN normalize-test
While not technically necessary, having normalized and log-transformed test expression values is helpful for visualization, so we'll do just that.

```{r}
test <- scrapper::normalizeRnaCounts.se(
    test,
    assay = <%= ASSAY %>,
:BEGIN block
    block = SummarizedExperiment::colData(test)[,<%= BLOCK %>],
:END
    num.threads = <%= NUM_THREADS %>
)
```
:END

# Reference data loading

We load one or more reference datasets to assign labels to each cell in our test dataset. 

```{r}
references <- list()
```

:BEGIN ref-body
## Reference `<%= SHORT_NAME_RAW %>`

:BEGIN celldex
We load our reference dataset from the `r BiocStyle::Biocpkg("celldex")` package:

```{r}
ref <- celldex::<%= CELLDEX_CMD %>(<%= CELLDEX_ARGS %>)
ref
```
:END

:BEGIN custom
We load our reference dataset:

```{r}
:BEGIN ref-setup
:END
ref
```
:END

:BEGIN replace-rownames
We replace the row names to obtain gene identifiers that are consistent with those used in the test dataset.

```{r}
new.rownames <- SummarizedExperiment::rowData(ref)[,<%= ALTERNATIVE_NAMES %>]
new.rownames[is.na(new.rownames)] <- ""
rownames(ref) <- new.rownames
head(new.rownames)
```
:END

Let's inspect the distribution of reference labels:

```{r}
table(SummarizedExperiment::colData(ref)[[<%= LABEL_FIELD %>]])
```

:BEGIN normalize
For proper marker detection, the reference expression values are expected to be normalized and log-transformed.

```{r}
ref <- scrapper::normalizeRnaCounts.se(
    ref,
:BEGIN block
    block = SummarizedExperiment::colData(ref)[,<%= BLOCK %>],
:END
    assay = <%= ASSAY %>,
    num.threads = <%= NUM_THREADS %>
)
```
:END

Now adding it to our reference list.

```{r}
references[[<%= SHORT_NAME %>]] <- ref
```
:END

:BEGIN intersect-ref
## Finding the intersection

For simplicity, we'll only consider the intersection of genes across all the test and reference datasets.
This improves the interpretability of correlations across references as they are computed from a common set of features.

```{r}
universe <- Reduce(intersect, lapply(references, rownames))
universe <- intersect(universe, rownames(test))
length(universe)
```
:END

# Annotation

We annotate the test cells with the `r BiocStyle::Biocpkg("SingleR")` package.
For each test cell, we compute Spearman correlations against the reference profiles.
The assigned label for each test cell is the one with the largest positive correlation.
For closely related labels, we recompute the correlations with the relevant subset of markers to improve resolution.
We do this for each reference separately and store the results in our list:

```{r}
all.trained <- list()
predictions <- list()
```

:BEGIN annotate-body
## Reference `<%= SHORT_NAME_RAW %>` {.tabset}

First we train a classifier from the reference data:

```{r}
ref <- references[[<%= SHORT_NAME %>]]
trained <- SingleR::trainSingleR(
    ref,
    labels = SummarizedExperiment::colData(ref)[[<%= LABEL_FIELD %>]],
    test.genes = rownames(test),
    assay.type = <%= ASSAY_NORM_REF %>,
    de.method = <%= DE_METHOD %>,<%= TRAIN_ARGS %>
    check.missing = FALSE,
    num.threads = <%= NUM_THREADS %>
)
all.trained[[<%= SHORT_NAME %>]] <- trained
```

And then we perform a prediction for each cell in the test dataset:

```{r}
pred <- SingleR::classifySingleR(
    test,
    trained,
    assay.type = <%= ASSAY_TEST %>,
    num.threads = <%= NUM_THREADS %>
)
predictions[[<%= SHORT_NAME %>]] <- pred
pred
```

We tabulate the assignments to examine the distribution of labels in the test dataset.

```{r}
table(pred$labels)
```

We visualize the (normalized) assignment scores for each test cell. 
Ideally, the highest scoring label should be unambiguous for each cell.

```{r}
SingleR::plotScoreHeatmap(pred, cells.order=pred$labels)
```

Another useful diagnostic is to examine the behavior of the top markers in the test dataset.
The cells assigned to a label are expected to upregulate that label's markers.
We should also check that the upregulated markers are biologically relevant for each label.

```{r, results="asis"}
for (lab in unique(pred$labels)) {
    cat("\n\n##", lab, "{-}\n")
    SingleR::plotMarkerHeatmap(
        pred,
        test,
        label = lab,
        assay.type = <%= ASSAY_NORM_TEST %>,
:BEGIN block
        block = SummarizedExperiment::colData(test)[,<%= BLOCK %>],
:END
:BEGIN old-row-names
        display.row.names = old.test.rownames,
:END
:BEGIN alt-row-names
        display.row.names = SummarizedExperiment::rowData(test)[,<%= DISPLAY_NAME_FIELD %>],
:END
        num.threads = <%= NUM_THREADS %> 
    )
}
```
:END

:BEGIN combining
## Combining predictions

We can combine the predictions from multiple references into a single set of results. 
For each test cell, we compute correlations to the predicted label from each reference, using the union of relevant markers from all references.
We then assign the label from the reference with the highest correlation, possibly after some fine-tuning. 

```{r}
combined <- SingleR::combineRecomputedResults(
    predictions,
    test,
    all.trained,
    assay.type.test = <%= ASSAY_TEST %>,
    num.threads = <%= NUM_THREADS %>
)
combined$orig.results <- NULL # strip these out to avoid redundancy.
combined
```

We can see how many test cells were assigned to each reference:

```{r}
table(names(predictions)[combined$reference])
```

As well as the variety of labels across all references:

```{r}
table(combined$labels)
```

Note that there might not be much diagnostic value in comparing the scores between references.
Multiple references may contain the same cell type, in which case ambiguous assignment of a single reference would be unavoidable.
:END

:BEGIN more-comparisons
# More comparisons {.tabset}

:BEGIN clustering
We tabulate the assigments with respect to pre-existing clustering results.
We expect to see some correlation between the clustering and assignments,
though major discrepancies may be of interest if they represent some novel factor of heterogeneity that does not exist in the reference.

```{r}
tab <- table(
    predicted = <%= TARGET %>$labels,
    cluster = SummarizedExperiment::colData(test)[,<%= CLUSTER_FIELD %>]
)
pheatmap::pheatmap(log2(tab+1), colors=colorRampPalette(c('white', 'blue'))(101))
```
:END

:BEGIN visualization
We plot the label assignments on pre-computed dimensionality reduction results.

```{r}
for (r in reducedDimNames(test)) {
    cat("\n\n##", r, "{-}\n")
    scater::plotReducedDim(test, r, colour_by = <%s= TARGET %>$labels)
}
```
:END
:END

# Saving results 

We create a directory in which to save results.

```{r save-directory}
result.dir <- "results"
dir.create(result.dir)
```

:BEGIN create-common-metadata
We also set up some extra metadata to add to each result.

```{r}
common.meta <- <%= COMMON_METADATA %>
```
:END

We then save our prediction results:

```{r save-pred}
:BEGIN save-body
anno.meta <- list(
    title='Cell type assignments from SingleR for reference `<%= SHORT_NAME_RAW %>`',
    authors = <%= AUTHOR %>,
    type="cell_label_assignment"
)
augere.core::saveResult(
    predictions[[<%= SHORT_NAME %>]],
    file.path(result.dir, <%= SAVE_PATH %>),
:BEGIN merge-metadata
    metadata=c(common.meta, anno.meta)
:END
:BEGIN no-merge-metadata
    metadata = anno.meta
:END
)
:END
```

:BEGIN save-combined
As well as the combined results.

```{r save-combined}
anno.meta <- list(
    title='Combined cell type assignments from SingleR', 
    authors = <%= AUTHOR %>,
    type="cell_label_assignment"
)
augere.core::saveResult(
    combined,
    file.path(result.dir, "combined"),
:BEGIN merge-metadata
    metadata=c(common.meta, anno.meta)
:END
:BEGIN no-merge-metadata
    metadata = anno.meta
:END
)
```
:END

# Session information {-}

```{r}
sessionInfo()
```
